Tackling irreproducible pathway enrichment analysis

### 2024-04-15 manuscript/manuscript.Rmd
updates

These observations highlight that pathway enrichment tools are essential in the interpretation of omics data.
However, there are concerns that inadvertent misuse of these methods leads to invalid results.
These potential problems include (i) the lack of a correct background for over-representation tests,
(ii) lack of p-value correction for multiple testing and (iii) general lack of reporting detail
[@Timmons2015-ex; @Reimand2019-uw; @Wijesooriya2022-li; @Zhao2023-ok].
Without comprehensive methodological documentation and reproducibility, it isn't possible to verify
the validity of the methods and results [@Sandve2013-lk].

To determine whether web based enrichment analysis is reproducible, here we systematically examine the
reproducibilty of DAVID enrichment analysis in 20 published articles, investigate root causes of
reproduction failure and describe some potential solutions.

@@

The results presented in the paper were compared to those we obtained, recording any
discrepancies.
Additionally, the statements made in the abstract, results, and discussion sections were carefully
examined and cross-checked with our results to understand whether those assertions survived reproduction
or not.

Each article was critically evaluated and given a score of one (1) to three (3).
Articles rated 1 showed overall poor reproducibility.
For example, less than 50% of GO and/or KEGG terms mentioned in the results and discussion were confirmed
upon reproduction, and the conclusions of the study were overall not supported.
Articles were rated 2 when at least 50% of the terms are observed as significant after replication
according to the original study's significance threshold, and most conclusions were supported.
Articles were rated 3 when they showed a high level of reproducibility, with 80% or more of the terms
shown in the article observed again in reproduction, and all conclusions were supported in reproduction.

@@

As reproducibility is closely associated with the quality of documentation, we collected key methodological
information about how the enrichment analysis was done and present in Table 1.
From this, we can see that only seven articles reported the version of DAVID that was used.
Version 6.7 was reportedly used in two articles while version 6.8 was specified in five.
Correction of p-values for multiple testing was specified or inferred in 11 articles, while the results of
the remaining nine articles did not take into consideration the importance of p-value correction for
minimising false positives.
Of the 20 articles, none mentioned explicitly that a custom background list was used for enrichment analysis.
We also examined whether the gene selection criteria were unambiguously clear, which was true in 16 articles.

@@

colnames(x) <- gsub("_"," ",colnames(x))
x |> gt(rownames_to_stub = TRUE) |> tab_header(title = "Table 1. Article reporting checklist results.")
#x %>% kbl(caption = "Table 1. Article reporting checklist results.") %>% kable_paper("hover", full_width = F)

@@

The R Markdown script called `example.Rmd` undertakes the transcriptome and enrichment analyses.
R Markdown allows the unification of extended descriptions, code and results into a single document,
which is becoming ubiquitous in bioinformatics and data science more generally
@Grolemund2017-zl.
The ability to provide extended descriptions is helpful for providing contextualising information about
the purpose of the work, detailed methods and interpretations derived from the analysis.
The R Markdown script contains code chunks that perform specific tasks including fetching an example
data set from DEE2, inspecting data quality, differential expression and two types of enrichment
analysis: over-representation analysis (ORA) and functional class scoring (FCS) (**Figure 3**).

@@

There are some notable limitations of this reproduction work.
The sample size of 20 is relatively small.
The articles were selected randomly from a list of articles that cited DAVID, and they are mostly from
low and mid-tier journals (mean SJR 2022: 1.2, SD=0.68) and the results may not be representative of
higher-tier journal articles.
This is not a complete reproduction of all work in those papers; we only reproduced the DAVID functional enrichment analysis,
which typically is just one component of an article that can have several other parts, including
experimental validation of key findings.
Therefore, if we identified that the enrichment analysis was not reproducible, it does not automatically mean that
all results and conclusions are flawed.

Furthermore, our reproducibility assessment is based on "head-counting" pathways that meet an arbitrary significance threshold.
This binary classification of results is less powerful as opposed to comparison of standardised effect sizes.

@@

@@

This article is an attempt to foster increased reproducibility in genomics by demonstrating the
implementation of the five pillars principles into a commonly used routine, namely differential
expression and pathway enrichment analysis.
However, these concepts may be difficult for novice analysts, and so we provide a step-by-step
protocol to guide them on how to customise template Dockerfile and R Markdown scripts to suit their
project's needs [@Ziemann2023-kz].
Although we aimed to make this process as easy as possible, it requires a substantial time outlay, which
we estimate at 15-20 hours.
This is significantly longer than using a web-based tool that provides results in just a few
minutes, but we consider the time investment is well spent.
It ensures analytical reproducibility, assists with auditing and makes the process of reproduction
simple and fast (Box 1), which should aid the early identification of errors.

@@

>>>>

These observations highlight that pathway enrichment tools are essential in the interpretation of omics data.
However, there are concerns that inadvertent misuse of these methods leads to invalid results
[@Timmons2015-ex; @Reimand2019-uw; @Wijesooriya2022-li; @Zhao2023-ok].
According to a recent survey of 186 articles, incorrect methods are very common [@Wijesooriya2022-li].
For example, a suitable background list for over-representation tests is reported in only 4.1% of studies.
Correction of p-values to account for multiple testing was reported in only 54% of studies.
This survey also revealed a general lack of reporting detail around other aspects of enrichment analysis,
such as the tool version used.

We hypothesise that the quality of enrichment analysis studies could be dramatically improved by adopting
established simple rules for reproducible computing [@Sandve2013-lk], and specific guidelines for
reporting enrichment analysis [@Zhao2023-ok].

Our aim is to determine how reproducible published enrichment analyses are.
To achieve this, we systematically examined the reproducibility of 20 published articles that used a web-based
enrichment tool.
We investigate root causes of reproduction failure and describe some potential solutions.

@@

The pathway enrichment results presented in the paper were compared to those we obtained.
It would be ideal to compare standardised effect sizes (fold enrichment scores) between original and
reproduction results, however these were reported in the main text of only 3/20 articles.
Therefore we adopted a binary classification system, based on head-counting pathways that were presented as
statistically significant in the original article.
We defined a pathway as confirmed when it reached a similar level of statistical significance
upon reproduction.
We define "similar significance level" to be a value within 50% of the value presented in the original article.
For example if an article shows a pathway with FDR=0.04, then an acceptable FDR in reproduction could be in
the range of 0.02 to 0.06.
Additionally, statements made in the abstract, results, and discussion sections based on specific pathway
enrichment results were carefully examined and cross-checked with our results to understand whether those
assertions survived reproduction or not.

Each article was critically evaluated and given a score of one (1; low) to three (3; high).
Articles were rated 1 when <50% of pathways shown in the results were confirmed
upon reproduction, and most conclusions derived from enrichment analysis were not supported by reproduction.
Articles were rated 2 when ≥50% of pathways shown in the results were confirmed
upon reproduction, and most conclusions derived from the pathway enrichment analysis were supported.
Articles were rated 3 when ≥80% of terms shown in the original article were observed again in
reproduction with a similar significance value, and all conclusions derived from the pathway analysis were
supported in reproduction.

@@

As reproducibility is closely associated with the quality of documentation, we collected key methodological
information about how the enrichment analysis was done and reported (Table 1).
Only seven articles reported the version of DAVID that was used.
Version 6.7 was reportedly used in two articles while version 6.8 was specified in five.
Correction of p-values for multiple testing was specified or inferred in 11 articles, while the results of
the remaining nine articles did not take into consideration the importance of p-value correction for
minimising false positives.
Of the 20 articles, none mentioned explicitly that a custom background list was used for enrichment analysis.
We also examined whether the gene selection criteria were unambiguously clear, which was true in 16 articles.

@@

colnames(x) <- gsub("_"," ",colnames(x))
x[,3]=NULL
x |> gt(rownames_to_stub = TRUE) |> tab_header(title = "Table 1. Article reporting checklist results.")
#x %>% kbl(caption = "Table 1. Article reporting checklist results.") %>% kable_paper("hover", full_width = F)

@@

The R Markdown script called `example.Rmd` undertakes the transcriptome and enrichment analyses.
R Markdown allows the unification of extended descriptions, code and results into a single document,
which is becoming ubiquitous in bioinformatics and data science more generally
@Grolemund2017-zl.
R Markdown supports languages other than R, including Python, Shell, Julia and others.
The ability to provide extended descriptions is helpful for providing contextualising information about
the purpose of the work, detailed methods and interpretations derived from the analysis.
The R Markdown script contains code chunks that perform specific tasks including fetching an example
data set from DEE2, inspecting data quality, differential expression and two types of enrichment
analysis: over-representation analysis (ORA) and functional class scoring (FCS) (**Figure 3**).

@@

There are some notable limitations of this reproduction work.
The sample size of 20 is relatively small.
The articles were selected randomly from a list of articles that cited DAVID, and they are mostly from
low and mid-tier journals (mean Scimago Journal Rank for 2022: 1.2, SD=0.68) and the results may not be representative of
higher-tier journal articles.
This is not a complete reproduction of all work in those papers; we only reproduced the DAVID functional enrichment analysis,
which typically is just one component of an article that can have several other parts, including
experimental validation of key findings.
Therefore, if we identified that the enrichment analysis was not reproducible, it does not automatically mean that
all results and conclusions are flawed.

Furthermore, our reproducibility assessment is based on "head-counting" pathways that meet an arbitrary significance range.
This binary classification of reproduction results is less powerful as opposed to comparison of standardised effect sizes [@Lakens2013-fx; @Goodman2016-ie].
For DAVID analysis, the fold enrichment score could be reported as a type of effect size.
Unfortunately, fold enrichment scores are infrequently reported, so it was not a feasible approach for a systematic and representative
reproduction study like this.
In future, we recommend researchers use both significance values and enrichment scores to prioritise and report
enrichment results.

@@

* Ambiguity around which tool was used, as some articles described using other tools in addition to DAVID
but did not label figures and tables with the tool used.

@@

This article is an attempt to foster increased reproducibility in genomics by demonstrating the
implementation of the five pillars principles into a commonly used routine, namely differential
expression and pathway enrichment analysis.
However, these concepts may be difficult for novice analysts, and so we provide a step-by-step
protocol to guide them on how to customise template Dockerfile and R Markdown scripts to suit their
project's needs [@Ziemann2023-kz].
Although we aimed to make this process as easy as possible, it requires a substantial time outlay, which
we estimate at 15-20 hours.
This is significantly longer than using a web-based tool that provides results in just a few
minutes, but we consider the time investment is well spent.
It ensures analytical reproducibility, assists with auditing and makes the process of reproduction
simple and fast (Box 1), which should aid the early identification of errors.
Using a literate scripting approach such as R Markdown, Jupyter or Quarto notebook, enables very high
levels of workflow transparency including inspection of intermediate data objects, and evidence that
the published methods were followed.

@@

Although container-based workflows are frequently described as being a complete solution to computational
reproducibility, they have some limitations.
For example, building a Docker image relies on a network of servers providing software packages which may work now
but not in a decade's time.
Therefore building such images has a short reproducibility horizon.
The Guix package and container manager, together with the Software Heritage code repository enables reproducible
image builds, which enhances the transparency of compute environments, and lengthening the reproducibility horizon
[@Vallet2022-gw].
It is also not guaranteed that Docker images published now will be compatible with future Docker versions,
whereas Guix can build an older version of itself.
